---
title: "Section2: Workbook 1"
author: "Chadi Ben Youssef"
output:
  html_document:
    df_print: paged
---
This workbook is adapted from Ch. 2 of your textbook and a previous iteration of this class (thanks to Dr Peter A. Krause).

Before starting, The document you will be working on is an RMarkdown notebook (similar to Jupyter for Python). This document allows to produce the same (HTML/DOC) file that we've seen in lecture.
It consists of sections where you can write good old regular text and other sections starting with "```{r}" where you can write code and run it.

To start, insert a code chunk below and run `rm(list=ls(all=TRUE))` inside it to make sure your global environment is clear to start.




Don't be afraid to edit the contents of chunks to follow the exercise instructions, to take notes (make sure use `#` for comments), to experiment with your own code ...!

### FUNCTIONS AND ARGUMENTS

R works using **FUNCTIONS**.  Functions carry out arithmetic operations, statistical operations, and basic bookkeeping/data management stuff.  

Functions take **ARGUMENTS** that tell them what to operate on and how. Find a function in R that computes the square root of a number.  How many **arguments** does this function take?

a. Try editing the code below to take the square roots of 5, 10, 125, 1e8.

```{r}
# Your code there 
# Note: notice the #, this means that this is a comment, not to be run as code, but helps understand the code.

```

### VECTORS
#### Creating & storing

Often, we want to store the output of a function.  This is typically stored as a **VECTOR**, an ordered sequence of values.  A vector can include a single value (in math, we'd call a single number a scalar, but R doesn't make any distinction).

When we wish to store output as a vector, we use the `<-` operator.  Note that we type the name of the new vector FIRST (to the left of the operator) and then define the function that will generate the output to the right of the operator.

a. Create a vector `sq_five` that contains the result of the square root of 5.

```{r}

```

The vector `sq_five`, now shows up in your **Global Environment** (typically in the upper-left in R Studio, but this can be changed in Tools-> Global Options-> Appearance). Go ahead and do it! Let's all have the same panel organization.

b. We can reveal the contents of a stored vector using the `print` function. Go ahead and 'print' the content of `sq_five`. 

```{r}

```

We can create a new vector using the function `c` (for concatenate or combine).  It takes a variable number of arguments, which together comprise the vector's elements (in the order given).

c. Create a vector `sequence` containing the numbers from 1 to 6:
  - using `c()`; 
  - using the operator `:`;
    and print the result.

```{r}

```

d. Create a vector called `odd_numbers` containing *odd* numbers from 1 to 111
  - using `c()`; (can you tho?) 
  - using a different function;
    and print the result.

```{r}

```

#### Sampling

We can *sample* elements from a vector using the `sample` function. It takes 4 arguments:

1. x: the vector to be sampled
2. size (integer): the size of the sample to be drawn
3. replace (bool): whether to replace one element before sampling the next one
4. prob: whether all elements have the same probability of being sampled.  (The NULL value to the "prob" argument makes all elements equally likely; we'll see down below how to change this.)

a. Sample 15 elements from `odd_numbers` with replacement. Do it 3 times and observe the results. What do you notice?

```{r}

```

b. If you simply want to reproduce a vector with its elements in a random order, you can use `sample` without any arguments except the vector itself. Try it:

```{r}

```

Let's say we want to make some elements of a vector more likely to be sampled than others.  We need to define a vector of sampling probabilities.  This vector *must be the same size as the one you're sampling from*.  The order of the probabilities lines up with the order of the elements in the vector you're sampling from.  The probabilities should add up to 1.0.  

I've made an example probability vector below. 

```{r}
probs <- c(0.05, 0.05, 0.1, 0.2, 0.15, .08, .25, .02, .06, .04)
```

c. You can verify whether an object in your Global Environment is a vector using the `is.vector` function. Try it:

```{r}

```

d. You can also determine how many elements are in a stored vector by using the `length` function.  try it:

```{r}

```

e. Now make sure that the sum of `probs` is equal to 1.0

```{r}

```
f. Now use `probs` to sample *without replacement* from the 10 last elements of `odd_numbers`:

```{r}

```

#### Types & combining

R allows vectors to include non-numeric elements.  In the example below, we have created a vector of text entries.  In R's mind, the entries in the vector below are of type `chr` as opposed to `num` (meaning they're string characters instead of numbers).  Elements of type `chr` are usually enclosed in quotes.

```{r}
some.names<-c("al", "bill", "chris")
print(some.names)
```

Note that the `c` (concatenate) function doesn't just operate on individual values, but also on whole vectors.  That is, you can combine vectors to make new ones!  

a. Create a vector `negative_odd` that contains the negative odd numbers from -111 to -1 and combine it with `odd_numbers` and store the result in a vector called `combined`.

```{r}

```

Let's say you want to "splice" a new value into an existing vector, and you don't just want to tag it onto the beginning or the end (which would be easy enough to do with `c`).  You can use `append`, which takes 3 arguments:

1. x: the vector you're changing
2. values: the value(s) you are adding in 
3. after: the index of the original vector after which the new value should be added  

b. Append the `negative_odd` with the value `-100` after the value `-101`. Make sure to  print the vector and check whether the change has been made. If not, why didn't it happen?

```{r}

```

R requires that all elements of a vector be of the same type (either `chr` or `num`). You can use the function `str` (for "structure") to report what type of elements a vector contains.

c. What happens to the `2` in the vector created by the code below?  Do you understand why?  

```{r}
mixture<-c("al", 2, "chris")
print(mixture)
str(mixture)
```

You can build new vectors out of repeating sequences, using the function `rep`.  
d. Given the vector `numbers`, construct a vector `repeated` that repeats the former 10 times.

```{r}
numbers<-c(1, 2, 3)

```

You can also construct a vector using the `seq` function (for "sequence").  It accepts 3 arguments: 

1. the starting value 
2. the ending value 
3. the size of the increment

e. What happens when you replace the `by` argument by `length.out`? 

```{r}
short <- seq(from=1, to=10, by=5)
long <- seq(from=1, to=10, by=50)

```


#### Modifying & subsetting

If your vector includes decimal-valued elements, you can apply the various functions below to the whole vector.

The `round` function rounds to the number of decimal places given in the second argument.  The `floor` function rounds down to the nearest whole integer below. The `ceiling` function rounds up to the nearest whole integer above. The `trunc` function (for "truncate") simply lops off the decimal places.  

a. Given `decimals` below:
  - round all numbers to `1` decimal place
  - round to the nearest whole integer below
  - round to the nearest whole integer up
  - change decimals in a way that `trunc(decimals)` produces the same result as `floor()`

```{r}
decimals <- c(1.15, 2.37, -4.56, -6.73)

```

Any operation you can perform on a vector, you can perform on one element of that vector.  To pick out an element, give the element's position using a bracketed number immediately following the vector's name.  

b. Round to one decimal the second to last element of decimals.

```{r}

```

#### Evaluating

Various evaluative operators can be applied to the elements of a vector.  (The full list of evaluative operators are on p.73 of your textbook.)

One such operator is the `==` ("double equals") symbol, which determines which elements in a vector to the left of the operator *EQUAL* the value to the right of the operator.  

c. Using the `==` operator determine, which elements in the vector `check` are equal to 4.

```{r}
check <- c(10:1)

```

We can get similar information to that returned by the code above reported in slightly different ways.  

The `table` function can be used to summarize the number of `TRUE` and `FALSE` returns, but unlike simple use of the operator, it doesn't give information about position.  

The `which` function reports *WHICH* element positions return `TRUE`.

d. try experimenting with both functions using `check == 4`. What do you notice?

```{r}

```

e. The `unique` function tells you which distinct types appear in a vector of tokens. Consider the vector `randos` below:
  - How many unique elements does it contain?
  - What's the smallest value?

```{r}
randos <- c(5, 1, 12, 6, 5, 10, 1, 2, 5, 12, 6, 10, 12, 5, 6, 12, 12, 2, 10, 12, 1, 2, 2, 1)

```


#### Sorting

You can sort a vector by numerical order using `sort`.  Using R help (?sort), sort `jumbled` in ascending order, then in descending order.

```{r}
jumbled <- c(1, 3, 7, 9, 10, 4, 6, 5, 2, 8)

```

### DATA FRAMES

Finally, let's work a bit with data frames.  Most datasets we will be working with in R are created in spreadsheet systems like Excel (we'll be working with .csv files or delimited .txt files). **Why should we avoid xls or xlsx files?**

You know what a spreadsheet looks like.  It has rows and columns.  Most of the time we'll just load them in from an outside source. 

However, it IS possible to create data frames in R itself.  Although this isn't a very realistic scenario, as an exercise it's good for revealing what's going on in R's internal workings.  To R, a data frame is a concatenation of several vectors (one for each column of the spreadsheet).

In the example below, we create several individual vectors (columns) and then use the "data.frame" function to stitch them together in a fixed order.  Run the code to see what the results look like.

Consider data frame `df`below, print a summary of the data frame.

```{r}
POS <- c("adj", "adv", "n", "conj", "prep")
TOKENFREQ <- c(421, 337, 1411, 458, 455)
TYPEFREQ <- c(271, 103, 735, 18, 37)
CLASS <- c("open", "open", "open", "closed", "closed")
df <- data.frame(POS, TOKENFREQ, TYPEFREQ, CLASS)
print(df)

```
Notice anything that needs changing?

Now let's load an existing DF that we saw in class. The file `_input/dataframe4.csv` contains data for the VERB into VERBing construction in the BNC (e.g., He [~V1~ forced] him into [~V2~ speaking] about it). For each instance of one such construction, the file contains

 - a column called `BNC`: the file where the instance was found (A06 in the first case);
 - a column called `VERB_LEMMA`: the lemma of the finite verb (force);
 - a column called `ING_FORM`: the gerund (speaking);
 - a column called `ING_LEMMA`: the lemma of the gerund (speak);
 - a column called `ING_TAG`: the part-of-speech tag of the gerund (VVG in the first case).

Load this file into a data frame `COV` and display the first six rows of `COV`.

```{r}

```

a. Delete the column with the corpus files; the new data frame is to be called `COV.2`.

```{r}

```

b. Delete the rows with the four rarest tags; the new data frame is to be called `COV.3`.

```{r}

```

c. From `COV.3`, create a new data frame `COV.4` which is sorted according to, first, the column `VERB_LEMMA` (ascending) and, second, the `ING_LEMMA` (descending).

```{r} 

```

d. Save the changed data frame into a text file `_output/dataframe5.csv`; use tab stops as separators, newlines as line breaks, and make sure you *don't have* row numbers and *no quotes*.

```{r}

```


   
